Attrition in Online Learning: Understanding Persistence and Dropout in Massive Open Online Courses

### 2014-10-27 paper_v1.tex
some updates to s1

\begin{abstract}
Persistence and attrition in MOOCs are systematically analyzed using self-report and behavioral data collected from [N] online learners in [M] courses. Study 1 offers insights into reasons for disengaging from MOOCs and explores relationships with prior behavior and reported intentions. Study 2 is a case study to develop a deeper understanding of attrition in MOOCs by conducting a case study. Targeting online learners who were predicted likely dropouts in a particular course were invited to provide feedback via a survey. (abstract needs work...)
\end{abstract}

@@

TODO: Outline different courses and go through variabels in the course info table.

Table:  course name; enrollment; pre-survey response rate; prop. demographics (age, gender, educ); who intended to do all; post-survey response rate; self-identified dropout rate

@@

Table: proportions for each dropout reason for each course

Table: correlations between reasons and one with cors between reasons and intentions

Plot: ecdf curves for proportion videos watched for each disengagement reason

@@

>>>>

\begin{abstract}
% Persistence and attrition in MOOCs are systematically analyzed using self-report and behavioral data collected from 20879 online learners in 18 courses. Study 1 offers insights into reasons for disengaging from MOOCs and explores relationships with prior behavior and reported intentions. Study 2 is a case study to develop a deeper understanding of attrition in MOOCs by conducting a case study. Targeting online learners who were predicted likely dropouts in a particular course were invited to provide feedback via a survey. (abstract needs work...)
Abstract goes here.
\end{abstract}

@@

Seventeen MOOCs on topics in a variety of disciplines were selected for investigation (see Table \ref{tab:s1sum}). For each course, we merged data from the initial course survey (distributed in the first weeks of the course), the feedback surveys (distributed in the last weeks of the course), and behavioral data collected in the learning environment. 
% Only those learners who completed the feedback survey were included in the analyses (some did not complete the initial course survey). 
Table \ref{tab:s1sum} provides summary statistics for enrollment, survey response rates, demographics, self-reported dropout, and behavioral indicators for each course. Self-reported dropout was measured in the feedback survey by asking learners if they stopped participating in the course. The three behavioral indicators are the proportions of learners who viewed at least 30, 50, and 80\% of the lecture videos. Information about age, gender, and education level were only available for learners who completed the initial course survey.

%Table:  course name; enrollment; pre-survey response rate; prop. demographics (age, gender, educ); who intended to do all; post-survey response rate; self-identified dropout rate

\subsubsection{Measures}

Learners reported reasons for enrolling in the course (using the OLEI scale \cite{kizilcec2015motivation} on the course survey) and leaving the course (on the feedback survey) on a grid of binary-choice items. The feedback survey for asked learners whether or not they have stopped participating before the end of the course, and if they did, what reasons influenced their decision to stop. For each reason, the student was asked to select one of two options: ``Did influence my decision to stop" and ``Did NOT influence my decision to stop". Compared to a question where the student only checks a box if the reason applies, this design aims to distinguish between respondents who skip reporting reasons and those who express that the reason did not apply to them. We also computed the fraction of videos in the course each learner had watched to relate self-report measures to learner behavior.

We first report the proportion for each reason for dropping out. This was computed as the number of respondents who reported that the reason applied to them divided by the total number of those who responded to the question (reporting that it applies or that it does not apply). We also computed the median and interquartile range of the proprtions across all courses for each reason. The resulting proportions are shown in Figure \ref{fig:prop-reason}. Second, we performed three correlational analyses were performed to investigate associations between these measures.

We examined the strength of correlations between reasons for dropout. The correlation matrix for the reasons were computed separately for each course, and then linearly combined to yield the average correlation matrix shown in Table \ref{tab:study1_rfd_corr}. Following the same procedure, we computed the correlation matrix between reasons to enroll and reasons for attrition. Furthermore, we computed correlations between reported reasons for dropout and certain learner behaviors. Finally, for each learner, we computed the fraction of course videos visited (the learner started playing the video) and fraction watched (the learner stayed with the video for 3 minutes, or for the whole video duration if shorter than 3 minutes). We then computed the matrix of correlations between reasons for dropout and these variables for each course. These matrices were averaged and the result is shown in Table [TableToBeAdded].

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{figure*}[ht]

\includegraphics[width=\maxwidth]{figure/prop-reason} \caption[Proportion of learners who indicated each reason for attrition across 17 MOOCs]{Proportion of learners who indicated each reason for attrition across 17 MOOCs\label{fig:prop-reason}}
\end{figure*}

\end{knitrout}

\begin{table*}
\label{tab:study1_rfd_corr}
\caption{Correlations among reasons for disengagement based on a linear combination of the courses' correlation matrices}
\small
\center
\begin{tabular}{lrrrrrrrrrrrrrr}
\toprule
 & 1 & 2 & 3 & 4 & 5 & 6 & 7 & 8 & 9 & 10 & 11 & 12 & 13 & 14 \\
\midrule
R1: Assn or exam was too difficult & 1.0 & 0.17 & 0.28 & 0.3 & 0.16 & 0.14 & 0.37 & 0.24 & 0.26 & 0.2 & 0.09 & 0.33 & 0.4 & 0.28 \\
R2: Couldn't keep up with deadlines &  & 1.0 & 0.06 & -0.03 & 0.01 & -0.05 & 0.05 & -0.07 & -0.06 & 0.32 & 0.17 & 0.04 & 0.08 & 0.04 \\
R3: Credentials not valuable to me &  &  & 1.0 & 0.34 & 0.34 & 0.25 & 0.34 & 0.37 & 0.39 & 0.17 & 0.17 & 0.27 & 0.28 & 0.35 \\
R4: Didn't enjoy online course format &  &  &  & 1.0 & 0.24 & 0.15 & 0.53 & 0.36 & 0.44 & 0.02 & 0.18 & 0.34 & 0.33 & 0.54 \\
R5: I was only exploring &  &  &  &  & 1.0 & 0.36 & 0.2 & 0.26 & 0.31 & 0.11 & 0.18 & 0.21 & 0.16 & 0.18 \\
R6: I learned all I wanted &  &  &  &  &  & 1.0 & 0.12 & 0.31 & 0.28 & 0.0 & 0.12 & 0.16 & 0.16 & 0.09 \\
R7: Material not explained properly &  &  &  &  &  &  & 1.0 & 0.36 & 0.45 & 0.15 & 0.13 & 0.31 & 0.35 & 0.66 \\
R8: Not challenging enough &  &  &  &  &  &  &  & 1.0 & 0.45 & 0.0 & 0.12 & 0.28 & 0.21 & 0.32 \\
R9: Not what I was looking for &  &  &  &  &  &  &  &  & 1.0 & 0.07 & 0.12 & 0.3 & 0.33 & 0.44 \\
R10: Course requires too much time &  &  &  &  &  &  &  &  &  & 1.0 & 0.06 & 0.08 & 0.19 & 0.12 \\
R11: I joined too late &  &  &  &  &  &  &  &  &  &  & 1.0 & 0.22 & 0.11 & 0.14 \\
R12: Technical difficulties &  &  &  &  &  &  &  &  &  &  &  & 1.0 & 0.27 & 0.3 \\
R13: Too advanced for my background &  &  &  &  &  &  &  &  &  &  &  &  & 1.0 & 0.33 \\
R14: Videos were confusing &  &  &  &  &  &  &  &  &  &  &  &  &  & 1 \\
\bottomrule
\end{tabular}
\end{table*}

\begin{table*}
\label{tab:study1_rfd_ei_corr}
\caption{Correlations between reasons for disengagement and enrollment intentions based on a linear combination of the courses' correlation matrices}
\small
\center
\begin{tabular}{lrrrrrrrrrrrrrr}
\toprule
 & R1 & R2 & R3 & R4 & R5 & R6 & R7 & R8 & R9 & R10 & R11 & R12 & R13 & R14 \\
\midrule
Career change & 0.03 & 0.02 & -0.11 & -0.03 & -0.02 & -0.06 & 0 & -0.01 & 0 & 0 & 0.03 & 0.04 & 0 & -0.03 \\
Experience online course & 0.06 & 0.01 & 0.01 & 0.1 & 0.17 & 0.09 & 0.02 & 0.01 & -0.01 & -0.02 & 0.05 & 0.12 & 0.12 & 0.08 \\
Fun and challenege & 0.05 & 0.07 & -0.03 & -0.01 & 0.02 & -0.05 & -0.01 & 0 & -0.08 & 0.06 & 0.02 & 0.01 & -0.01 & 0 \\
General interest & 0.01 & -0.04 & -0.04 & 0.02 & -0.01 & -0.07 & -0.05 & 0.01 & 0.04 & 0.03 & 0.06 & -0.03 & -0.01 & 0.02 \\
Improve my English & 0.08 & 0 & 0.05 & 0.1 & 0.09 & 0.07 & 0.02 & 0.03 & 0.04 & 0 & 0.1 & 0.17 & 0.08 & 0.05 \\
Presitgious uni/prof & 0.03 & 0.15 & -0.08 & -0.07 & -0.04 & -0.06 & -0.07 & -0.1 & -0.14 & 0.03 & 0.02 & 0 & -0.03 & -0.05 \\
Relevant to job & -0.04 & -0.01 & 0.02 & 0.05 & 0.04 & 0 & 0.01 & -0.02 & 0 & 0 & -0.03 & 0.09 & 0.02 & 0.01 \\
Meet new people & 0.05 & -0.02 & 0.05 & 0.08 & 0.06 & -0.04 & 0.03 & 0.13 & -0.01 & 0.01 & 0.14 & 0.15 & 0.06 & 0.03 \\
Other & 0 & -0.07 & 0.08 & 0.05 & 0 & -0.03 & 0.03 & 0 & 0.01 & -0.08 & 0 & 0.09 & 0.15 & 0 \\
Personal growth & 0.02 & 0.16 & 0 & -0.03 & 0.02 & -0.02 & 0.07 & -0.01 & 0 & 0.03 & 0.1 & 0 & -0.03 & -0.02 \\
Relevant to research & -0.02 & 0.01 & 0.1 & 0.06 & 0.05 & 0 & 0.02 & 0.03 & -0.01 & -0.02 & 0.13 & 0.1 & -0.01 & 0.07 \\
Relevant to school & -0.06 & 0 & 0.08 & 0.01 & 0.08 & 0.02 & -0.01 & 0.02 & -0.05 & -0.04 & 0.08 & 0.02 & -0.02 & -0.01 \\
Statement of accomp. & 0.08 & 0.1 & 0.02 & 0.03 & 0.01 & -0.02 & 0.08 & 0 & 0.02 & 0 & 0.06 & 0.09 & 0.14 & 0.02 \\
Take with others & 0 & -0.02 & 0.16 & 0.06 & 0.12 & 0.05 & 0.06 & 0.24 & 0.11 & 0.04 & 0.09 & 0.15 & 0.2 & 0.08 \\
\bottomrule
\end{tabular}
\end{table*}

(Note: this section is work in progress, we will go through each RQ and present results that address them here.)

Table \ref{tab:study1_rfd_ei_corr}.

How do prior learner characteristics (demographics, geographic location, intentions and motivations, and previous online courses) influence learners' likelihood of dropping out of a MOOC? (RQ1)

What proportion of learners who dropped out were satisfied with their achievements in the course? (RQ2)

What reasons do learners report for dropping out of a MOOC? (RQ3a)

 How do reasons for dropping out vary across prior learner characteristics (cf. RQ1)? (RQ3b)

How does behavior in MOOCs vary by reasons for dropping out? (RQ4)

The two most commonly reported reasons for dropout are related to one's ability to commit time to the course. The correlation between the two reasons is $r=0.32$. A more detailed analysis we conditioned on responses to these two reasons revealed that 30\% of respondents who reported at least one of these two time reasons reported no other reasons for dropout, while 24\% reported one other reason and 46\% reported more than one.

Difficulty of an assignment or exam was the third most reported reason for dropout. It is most positively correlated with reasons citing difficulties, including ``Too advanced for my background'' ($r=0.4$), ``Materials not explained properly'' ($r=0.37$), and ``Technical difficulties'' ($r=0.33$).

The proportion of respondents who selected ``Technical difficulties'' exceeded 20\% in 10 out of the 17 courses. This reason, however, exhibits positive correlations stronger that 0.2 with 7 other reasons. [What might we like to study more about respondents who selected this reason?]

In the median, 19\% of respondents reported they learned what they wanted in the course, and this reason was moderately positively correlated with reasons citing no or incomplete interest in the course, namely ``I was only exploring the course'' ($r=0.36$), ``The course is not challenging enough for me'' ($r=0.31$), and ``The course is not what I was looking for'' ($r=0.28$).

Another set of interesting positive correlations exists between ``Course is not what I was looking for'' and other reasons citing the materials' level of difficulty/challenge, the value of the credentials to the respondent, and the extent to which she enjoyed the online course format, suggesting .... [TBC]

The reason that exhibited the strongest positive correlations with other features was ``Videos were confusing''. it correlates most strongly with ``Materials not explained properly'' ($r=0.66$) and ``Did not enjoy online course format'' ($r=0.54$)

Unlike correlations among reasons for disengagement, correlations of reasons for disengagement with enrollment intentions were mostly weak ($-0.14 < r < 0.24$). Among the strongest positive correlations were the cases listed in Table \ref{tab:rfd_ei_strongest_correlations}

% \begin{table*}
% \label{tab:rfd_ei_strongest_correlations}
% \caption{Strongest observed correlations ($r >= 0.15$) between reasons for disengagement and enrollment intentions}
% \center
% \small
% \begin{tabular}{llr}
% \toprule
%  Enrollment intention & Dropout reason & $r$ \\
% \midrule
% Take with others & Course not challenging enough & 0.24 \\
% Take with others & Too advanced for my background & 0.20 \\
% Experience an online course & I was only exploring the course & 0.17 \\
% Improve my English & Technical difficulties & 0.17 \\
% Personal growth & Could not keep up with deadlines & 0.16 \\
% Take with others & Credentials not valuable enough for me & 0.16 \\
% Take with others & Technical difficulties & 0.15 \\
% Prestigious uni/prof & Could not keep up with deadlines & 0.15 \\
% \bottomrule
% \end{tabular}
% \end{table*}

@@

(Note: this section is also work in progress.)

Even though study 1 accentuates ``time'' as a reported reason for dropout, interpreting what time means is not straight forward. Expectency-value theory models the learner as a rational agent who allocates time to different activities to maximize subjective gain. In other words, a learner might disengage from the course if she gets occupied by other higher priority tasks. Because learners can join or leave MOOCs freely. Joining a MOOC does not necessarily imply that a learner is capable of allocating the time for it according to the rational framework that expectancy value theory predicts. Moreover, previous research on volitional control suggests that learners allocate time to the next highest priority task, but could rather get distracted by lower priority activities that promise nearer gratification. So what does lack of time for the course mean? In study 2, we attempt to investigate various possible dropout reasons that learners might report as ``lack of time".
